Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-09-27 manuscript.Rmd
kväll 27e

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negative effect on vitals and the capacity of patients to tolerate certain interventions. This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results. 
<!-- (ÅTERKOM: har inte råd att refereas?) -->

@@

Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward. Because examining admission criteria to the ICU is part of the analysis it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals. Propensity score matching can be used to identify differences in ICU admission criteria between hospitals and find patients that were admitted but need not have been. These patients are excluded from predictor selection to reduce this bias. This is part of the development and training of the PSM model.

@@

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.

Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India. It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contains a series of steps and the work flow should understood as linear, but iterative. The output of each step is the input for later steps. Therefore the output of the last step contain the cumulative error from unfit assumptions made the second and fourth step. The output is tested against Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

<!-- ÅTERKOM inte klart bara gissat -->

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is a Training Part used to develop and train the PSM, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method described in chapter Bias. Training Data is used in steps 2-4 and the output is then validated using Validation Data.

@@

Third step is to build a standard multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables. This model is used to calculate the standard variable propensity score for all patients. The propensity score determines the likelihood that a patient will receive ICU care, given their condition.

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertently  vastly under- or overestimate the  need of ICU beds remain.  
The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one from Training Data. The proportion of this subset versus the subset that did receive ICU care is also the proportion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

>>>>

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher levels of malnutrition and comorbidities and would have a negative effect on vital parameters and the capacity of patients to tolerate both injuries and interventions. This circumstance doesn't affect the analysis and internal validity, but reduces the generalizability of the results.

@@

Hospitals with trauma wards could have lower admission criteria to the ICU than hospitals with no trauma ward. Because examining admission criteria to the ICU is part of the analysis, any bias from differences in hospital admission criteria must be minimized. This is why the PSM model is trained using a training and validation subset as described in the chapter Statistical methods. If this proves insufficient and unsatisfactory it may be necessary to examine and address this bias by comparing ICU admitted patients between hospitals using the same PSM methodology, but this will go beyond the time frame and scope of this thesis.

@@

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment. 
Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, that is indicative there exists shortage of ICU beds in tertiary hospitals in India.

The proportion of the subset of patients that should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that should hade received ICU care based on our prediction model but did not due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contains a series of steps. The output of each step is the input for later steps. The model is developed using Testing Data and is tested with Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is Training Data used to develop the propensity score model, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method. The prediction model produced in step 3 is based on and fitted against the Training data and predictors are weighed so that the the model awards patients that did go to the ICU corresponding high propensity. This model can then be used to predict new patients propensity to go to the ICU based on their baseline predictor values, and by saving 40% in the Validation data set we have a set of patients that is new to the model on which we may test how well the model predicted their fate. 
Training Data is used in steps 2-3 and the resulting model is then validated using Validation Data.

@@

Third step is to build a multivariable logistic regression model with ICU care as the outcome and the predictors selected in the second step as independent variables using Testing Data.

Fourth step is to apply this model on patients from the Validation Data set, and see whether the model correctly prediced which patients in the Validation data received ICU treatment. Using the model, the standard variable propensity score is calculated for each patient in the Validation Data set. If the patients that were admitted to the ICU in the validation set receive appropriate propensity scores by the model constructed in step 2-3, the model is valid. The propensity score determines the likelihood that a patient will receive ICU care, given their condition. 
<!-- 
#fråga: är det såhär man gör menade du Johanna? vad är det egentligen som händer i steg 3 och vad består "träningen" av i steg 2-3? År det att man har valt sina variabler i steg 2 men vi vet inte om de är bra och vi vet inte deras inbördes relevans/vikt för den oberoende variabeln. Så vi petar in dem i steg 3 och klickar OK, och sen bygger R en multivariabelregressionsmodell åt en som viktar prediktorerna åt en så att modellen tilldelar patienterna propensity scores som stämmer bäst överens med deras ICU-status (som är känt för R). R tweakar/vikar/manipulerar modellen att passa träningsdatasetet man matat den med, och bygger en modell som mha prediktorerna är bäst på att identifiera de patienter som R vet hamnade på ICU?
Det gör vi för att optimera modellens kapacitet att "förutse" vad som kommer hända med en "helt ny" patient med en uppsättning värden på prediktorerna? 
Så har vi i Validation Data sparat en uppsättning "helt nya" patienter som vi prövar modellen på? Och ser om modellen är bra på att ge ICU-patienter rimliga propensity scores?
Är det det du menar att modellen ska "träna" på ett set och testas på ett annat?
Och om man switchar över redan i steg 4 finns det ingen validering sen mera för dålig PSM i steg 5? -->

Fifth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable. There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria and inadvertently vastly under- or overestimate the  need of ICU beds remain.  
<!-- (b) Fråga: 1. visst finns dessa svagheter i propensity score matching?
2. vad gör man åt dem? 
3. är beskrivningen av åtgärder mot detta något som hör hemma här eller i results?-->

The proportion of the subset of patients that probably should hade received ICU care based on our prediction model versus the subset that did receive ICU care is also the proportion of ICU bed shortage.
